---
title: "Practical 5"
authors: Eleonora Mazzocca, Nùria Torquet
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    # df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

# PART 1 - Practice 12 R: ARFIMA Model

## 1.1 Simulating the ARFIMA Model

In this section we will generate a time series using an ARFIMA(2,d,1)
model. We expect "long memory" in data through the parameter $d$, which
we for now choose as $d=0.4$. We set a seed so that the same simulated
series, and therefore the same estimates, are reproduced when the
document is knitted again.

```{R}
set.seed(1)
n <- 1000
x <- arima.sim(n, model = list(phi = c(0.2, 0.1), dfrac = 0.4, theta = 0.9))
plot(x, type = "l", main = "Simulated ARFIMA(2, 0.4, 1)", ylab = "Value", xlab = "Time", col = "steelblue")
```

## 1.2 Fitting the ARFIMA Model

Now we will fit the best model. To do so, we first install the package
*fracdiff* which we need to estimate the best parameters. Then we use
*fracdiff* with the following parameters:

-   **nar=2:** specifies 2 Auto-Regressive (AR) terms

-   **nma=1**: specifies 1 Moving Average (MA) term

-   **drange=c(0,0.5)**: restricts the search for $d$ to the range (0 to
    0.5)

```{R}
# install.packages("fracdiff")
library(fracdiff)

fit1<-fracdiff(x, nar=2, nma=1, drange=c(0,0.5))
summary(fit1)

# Extract d and remove the long-memory component
d<-fit1$d
fit2=diffseries(x,d)
acf(fit2, main = "ACF of Fractionally Differenced Series")

# Fit the remaining ARMA structure
fit3<-arima(fit2,order=c(2,0,1))
fit4<-resid(fit3)
acf(fit4, main = "ACF of Residuals (IID Noise Check)")


# Formal test for independence
print(Box.test(fit4, lag = floor(log(n)), type = "Ljung-Box"))
```

The residual diagnostics are compatible with white noise after removing
the estimated long-memory and ARMA components:

-   the Ljung-Box test does not reject independence at the $5\%$ level,
    so there is no strong evidence of remaining serial correlation in
    `fit4`

-   however, the parameter recovery is poor: in this simulated example
    the estimated $d$ is far from the true value $0.4$, which is exactly
    the practical issue highlighted in the guide for more complex ARFIMA
    models

**Observations on ACF plots:**

-   **ACF of Fractionally Differenced Series:** After applying
    `diffseries(x, d)`, the strong "long-memory" (slowly decaying)
    correlations should be mostly removed. Any remaining dependence
    should now be short-range and attributable to the ARMA part. In this
    run the ACF does not show a very clear residual pattern, which
    illustrates how difficult it can be to separate long memory from
    short-memory dynamics in finite samples.

-   **ACF of Residuals (`fit4`):** Once the $ARMA(2,1)$ structure is
    removed from the fractionally differenced series, the ACF shows no
    dominant spikes outside the confidence bounds. This suggests the
    residuals behave approximately like white noise.

## 1.3 Change different values for d, phi, theta, n

Let's see how changing parameters affects the model's ability to recover
$d$.

In particular, we analyze the influence of the sample size, the
fractional differencing parameter d, and the AR/MA coefficients.

### Scenario A: High Data Volume ($n=5000$) with Stronger $AR$

First, we increase $n$ to $5000$ and strengthen the first $AR$
coefficient (while weakening the $MA$ one) to see if more data helps the
model separate short-term and long-term memory.

```{r}
set.seed(1)
n_a <- 5000
x_a <- arima.sim(n_a, model = list(phi = c(0.6, 0.1), dfrac = 0.3, theta = 0.5))

plot(x_a, type = "l", main = "ARFIMA: n=5000, d=0.3, Strong AR", col = "darkgreen")
```

```{r}
library(fracdiff)

fit1<-fracdiff(x_a, nar=2, nma=1, drange=c(0,0.5))
summary(fit1)

# Extract d and remove the long-memory component
d<-fit1$d
fit2=diffseries(x_a,d)
acf(fit2, main = "ACF of Fractionally Differenced Series")

# Fit the remaining ARMA structure
fit3<-arima(fit2,order=c(2,0,1))
fit4<-resid(fit3)
acf(fit4, main = "ACF of Residuals (IID Noise Check)")


# Formal test for independence
print(Box.test(fit4, lag = floor(log(n_a)), type = "Ljung-Box"))
```

Even with many more observations, the estimate of $d$ is still very poor
in this run: it is pushed close to $0$ instead of the true $d=0.3$. This
shows that increasing $n$ does not automatically solve the
identification problem when the series contains both long memory and
non-trivial ARMA dynamics.

Even though the `fracdiff` function failed to correctly identify the
*source* of the memory (it missed $d$), the final combination of
fractional differencing and $ARIMA(2,0,1)$ was still mathematically
flexible enough to remove most of the remaining dependence, as indicated
by the residual ACF and the Ljung-Box test.

### Scenario B: Low Data ($n=500$) with "Boundary" Parameters

Here we reduce $n$ and set $d$ close to the non-stationary boundary
($0.45$) and a high $MA$ term ($0.9$).

```{r}
set.seed(42)
n_b <- 500
x_b <- arima.sim(n_b, model = list(phi = c(0.1, 0.1), dfrac = 0.45, theta = 0.9))

plot(x_b, type = "l", main = "ARFIMA: n=500, d=0.45, High MA", col = "darkred")
```

```{r}
library(fracdiff)

fit1<-fracdiff(x_b, nar=2, nma=1, drange=c(0,0.5))
summary(fit1)

# Extract d and remove the long-memory component
d<-fit1$d
fit2=diffseries(x_b,d)
acf(fit2, main = "ACF of Fractionally Differenced Series")

# Fit the remaining ARMA structure
fit3<-arima(fit2,order=c(2,0,1))
fit4<-resid(fit3)
acf(fit4, main = "ACF of Residuals (IID Noise Check)")


# Formal test for independence
print(Box.test(fit4, lag = floor(log(n_b)), type = "Ljung-Box"))
```

Also in this case the parameter $d$ is not estimated correctly. This is
coherent with what noted in the practical guide: adding $MA$ components
(like $\theta=0.9$) makes things "worse" because the short-term shock of
the $MA$ term can perfectly mimic or mask the long-memory effects of $d$
in small samples. Here the problem is amplified because $d=0.45$ is
already close to the stationarity boundary, so distinguishing persistent
long memory from strong short-run dependence becomes even harder.

## 1.4 Simpler model: ARFIMA(1,d,0)

We now analyze a simpler ARFIMA(1,d,0) model in order to compare the
estimation performance with the previous example.

```{r}
library(fracdiff)
set.seed(42)
n<-1000
sample<-fracdiff.sim(n, ar=0.9, d=0.4)
x<-sample$series
plot(x,type="l")
fit1<-fracdiff(x,nar=1)
summary(fit1)
d<-fit1$d
y1<-diffseries(x,d)
acf(y1)
y2<-arima(y1,order=c(1,0,0))
z<-resid(y2)
acf(z)
Box.test(z,lag=floor(log(n)),type=c("Ljung-Box"))
```

In this simpler configuration, the estimated parameters are much closer
to the true simulated values. The estimated $d$ is near $0.4$ and the
estimated AR coefficient is near $0.9$, so the model recovers both the
long-memory and short-memory parts substantially better than in the
previous ARFIMA$(2,d,1)$ examples.

This suggests that ARFIMA estimation performs better when the underlying
model structure is less complex.

Now we will see what happens if we change to an ARFIMA$(0,d,1)$ model.

## 1.5 ARFIMA(0,d,1) model

We keep the same idea as in the previous section, but now the
short-memory component is purely moving-average. This lets us compare
whether estimation remains reasonably accurate when the model has only
one MA parameter together with fractional differencing.

```{r}
library(fracdiff)
set.seed(3)
n <- 1000
sample <- fracdiff.sim(n, ma = 0.9, d = 0.4)
x <- sample$series
plot(x, type = "l", main = "Simulated ARFIMA(0, 0.4, 1)")

fit1 <- fracdiff(x, nma = 1, drange = c(0,0.5))
summary(fit1)

d <- fit1$d
y1 <- diffseries(x, d)
acf(y1, main = "ACF after Fractional Differencing")

y2 <- arima(y1, order = c(0,0,1))
z <- resid(y2)
acf(z, main = "ACF of Residuals")

Box.test(z, lag = floor(log(n)), type = "Ljung-Box")
```

In this case the estimation is again clearly better than in the
ARFIMA$(2,d,1)$ examples. The estimated $d$ is close to the true value
$0.4$, and the estimated MA coefficient is also close to the simulated
$\theta = 0.9$.

After removing the fractional dependence, the ACF of `y1` still shows
the short-run pattern expected from an MA component, and after fitting
the MA$(1)$ model the residual ACF no longer shows important
autocorrelation. The Ljung-Box test is also non-significant, so the
residuals are compatible with white noise.

Together with Section 1.4, this confirms the main message of the
practical: ARFIMA estimation tends to work better for simpler model
structures, whereas richer ARMA dynamics can hide the true fractional
parameter $d$.

## 1.6 Real data: Australian residents

We now apply ARFIMA modeling techniques to a real dataset. The `austres`
dataset contains quarterly estimates of the Australian resident
population between 1972 and 1993.

```{r}
library("datasets")
x<-austres
acf(x, main = "ACF of Australian residents (raw)")

# First difference to handle trend
dx<-diff(x)
acf(dx, main = "ACF of differenced Australian residents")

# Fit fractional d to the differenced series
fdx<-fracdiff(dx,nar=0,nma=0, drange=c(0,0.5))
summary(fdx)
d=fdx$d
d

# Remove fractional noise
resid<-diffseries(dx,d)
y<-resid-mean(resid)

acf(y, main = "Final ACF: white noise check")
Box.test(y,lag=floor(log(length(x))),type=c("Ljung-Box"))
```

**Observations:**

-   **ACF of Raw Data:** shows extremely slow linear decay, indicating
    the series is **non-stationary** and has a strong trend.

-   **ACF of Differenced Data:** after `diff(x)`, the trend is removed,
    but significant autocorrelations remain at seasonal lags, which
    shows that the series still has serial dependence.

-   **Final ACF:** after removing the estimated fractional $d$
    component, the remaining autocorrelations are much smaller, but the
    plot can still show some moderate spikes, especially around seasonal
    lags. So the result is closer to white noise than the raw series,
    but not perfectly featureless.

The **Ljung-Box test** on `y` does not reject independence at the $5\%$
level, so this transformation removes a large part of the dependence.
However, the estimated fractional parameter is not tiny, and the final
ACF suggests that some structure may still remain. A cautious conclusion
is therefore that first differencing plus fractional differencing gives
a much more stationary series, but not necessarily a perfect white noise
process.

Overall, these results show that ARFIMA models are useful for capturing
long-memory behavior, but their estimation is clearly more reliable in
simpler specifications than in more complex ARMA settings.
