---
title: "Practical 2 — Testing IID and Gaussian Characters"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# Practical 2A

## 1. Generate samples of lenght n of different IID noises using the following instructions

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
n <- 1000

x1 <- rnorm(n, mean = 0, sd = 1)
x2 <- runif(n, min = 0, max = 1)
x3 <- rexp(n, rate = 2)
x4 <- rchisq(n, df = 3)
x5 <- rpois(n, lambda = 4)
x6 <- rt(n, df = 5)

plot(x1, type="l")
hist(x1, 30)

plot(x2, type="l")
hist(x2, 30)
```

Observe what happens when you move the parameters. Try different distributions and parameters. What do you observe?

### Normal distribution: change mean and sd

```{r}
n <- 1000

x1 <- rnorm(n, mean = 0, sd = 1)
x2 <- rnorm(n, mean = 5, sd = 1)
x3 <- rnorm(n, mean = 0, sd = 3)

par(mfrow=c(3,2))

hist(x1, 30, main="Normal mean=0 sd=1")
plot(x1, type="l")

hist(x2, 30, main="Normal mean=5 sd=1")
plot(x2, type="l")

hist(x3, 30, main="Normal mean=0 sd=3")
plot(x3, type="l")
```

We see that:
- Increasing the mean shift the distribution
- Increasing sd spreads the data

### Expoential distribution: change rate

```{r}
x1 <- rexp(n, rate=0.5)
x2 <- rexp(n, rate=3)
x3 <- rexp(n, rate=10)

par(mfrow=c(3,1))

hist(x1, 30, main="rate=0.5")
hist(x2, 30, main="rate=1")
hist(x3, 30, main="rate=3")
```

We see that:
- Higher rate implies that values are more concentrated around 0
- Lower rate implies that values are more spread out, longer

### Uniform distribution: change rate

```{r}
x1 <- runif(n, min=0, max=1)
x2 <- runif(n, min=-5, max=5)
x3 <- runif(n, min=10, max=20)

par(mfrow=c(3,1))

hist(x1, 30, main="0 to 1")
hist(x2, 30, main="-5 to 5")
hist(x3, 30, main="10 to 20")
```

We see that the distribution shape is the same, but the values are shifted and more spread out when we increase the range.

### Chi-square: change the degrees of freedom

```{r}
x1 <- rchisq(n, df=1)
x2 <- rchisq(n, df=5)
x3 <- rchisq(n, df=15)

par(mfrow=c(3,1))

hist(x1, 30, main="df=1")
hist(x2, 30, main="df=5")
hist(x3, 30, main="df=15")
```

We see that:
- Small df implies that values are more concentrated around 0, with a long tail
- Large df implies that values are more spread out, with a shorter tail

### Poisson distribution: change lambda

```{r}
x1 <- rpois(n, lambda=1)
x2 <- rpois(n, lambda=5)
x3 <- rpois(n, lambda=15)

par(mfrow=c(3,1))

hist(x1, 30, main="lambda=1")
hist(x2, 30, main="lambda=5")
hist(x3, 30, main="lambda=15")
```

We see that:
- Small lambda implies that values are more concentrated around 0, with a long tail
- Large lambda implies that values are more spread out, with a shorter tail

### t distribution: change the degrees of freedom
```{r}
x1 <- rt(n, df=1)
x2 <- rt(n, df=5)
x3 <- rt(n, df=30)

par(mfrow=c(3,1))

hist(x1, 30, main="df=1")
hist(x2, 30, main="df=5")
hist(x3, 30, main="df=30")
```

We see that:
- Small df implies that values are more concentrated around 0, with a long tail
- Large df implies that values are more spread out, with a shorter tail, and when df is large the distribution looks more like a normal distribution

Conclusion: moving the parameters change
- location -> mean
- spread -> variance/sd
- skewness -> rate/df
- tail -> rate/df

Observe what happens when you move the parameters.
For different distributions, do for example

```{r}
n <- 1000
k <- 30
x <- rexp(n, rate=2)
plot(x, type="l")
hist(x, k)
```

Here k denotes the number of classes for the histogram.
Test with other distributions writing distributions in the help utility.

To explore all distributions in R:
```{r}
help("Distributions")
```

or 
```{r}
?rexp
?rnorm
```


## 2. Verify the IID character of any sample of the previous exercise. Use Ljung-Box test with the instruction
```{r}
Box.test (x, lag=h, type="Ljung")
```

For example:
```{r}
n <- 10000
k <- 3
x <- rt(n, df=k)
plot(x, type="l")
Box.test(x, lag=h, type="Ljung")
```

Interpretation:
We see a large p-value, which means that we cannot reject the null hypothesis of independence, so we can conclude that the data are independent.
The sample behaves like IID noise, as expected.

We can try different lags, from h=1, to log(n):
```{r}
for(h in 1:log(n)){
  print(Box.test(x, lag=h, type="Ljung"))
}
```
We see that for all lags, we cannot reject the null hypothesis of independence, so we can conclude that the data are independent.
p-values are all large, confirming that the data are independent.

### Comparison with random walk

A random walk is not IID because each value depends on the previous one. We can verify this by applying the Ljung-Box test to a random walk:
```{r}
n <- 10000
rw <- cumsum(rnorm(n))
plot(rw, type="l")
Box.test(rw, lag=log(n), type="Ljung")
```
We see a small p-value, which means that we reject the null hypothesis of independence, so we can conclude that the data are not independent.
The random walk does not behave like IID noise, as expected.

We can try different lags for the random walk as well:
```{r}
for(h in 1:log(n)){
  print(Box.test(rw, lag=h, type="Ljung"))
}
```
We see that for all lags, we reject the null hypothesis of independence, so we can conclude that the data are not independent.
p-values are all small, confirming that the data are not independent.