---
title: "Practical 2 — Testing IID and Gaussian Characters"
authors: Eleonora Mazzocca, Nùria Torquet
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

# Practical 2A

## 1. Generate samples of length n of different IID noises using the following instructions

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(forecast)
```

```{r}
n <- 1000

x1 <- rnorm(n, mean = 0, sd = 1)
x2 <- runif(n, min = 0, max = 1)
x3 <- rexp(n, rate = 2)
x4 <- rchisq(n, df = 3)
x5 <- rpois(n, lambda = 4)
x6 <- rt(n, df = 5)

plot(x1, type="l")
hist(x1, 30)

plot(x2, type="l")
hist(x2, 30)
```

Observe what happens when you move the parameters. Try different
distributions and parameters. What do you observe?

### Normal distribution: change mean and sd

```{r}
n <- 1000

x1 <- rnorm(n, mean = 0, sd = 1)
x2 <- rnorm(n, mean = 5, sd = 1)
x3 <- rnorm(n, mean = 0, sd = 3)

par(mfrow=c(3,2))

hist(x1, 30, main="Normal mean=0 sd=1")
plot(x1, type="l")

hist(x2, 30, main="Normal mean=5 sd=1")
plot(x2, type="l")

hist(x3, 30, main="Normal mean=0 sd=3")
plot(x3, type="l")
```

We see that: - Increasing the mean shift the distribution - Increasing
sd spreads the data

### Exponential distribution: change rate

```{r}
x1 <- rexp(n, rate=0.5)
x2 <- rexp(n, rate=3)
x3 <- rexp(n, rate=10)

par(mfrow=c(3,1))

hist(x1, 30, main="rate=0.5")
hist(x2, 30, main="rate=1")
hist(x3, 30, main="rate=3")
```

We see that: - Higher rate implies that values are more concentrated
around 0 - Lower rate implies that values are more spread out, longer

### Uniform distribution: change rate

```{r}
x1 <- runif(n, min=0, max=1)
x2 <- runif(n, min=-5, max=5)
x3 <- runif(n, min=10, max=20)

par(mfrow=c(3,1))

hist(x1, 30, main="0 to 1")
hist(x2, 30, main="-5 to 5")
hist(x3, 30, main="10 to 20")
```

We see that the distribution shape is the same, but the values are
shifted and more spread out when we increase the range.

### Chi-square: change the degrees of freedom

```{r}
x1 <- rchisq(n, df=1)
x2 <- rchisq(n, df=5)
x3 <- rchisq(n, df=15)

par(mfrow=c(3,1))

hist(x1, 30, main="df=1")
hist(x2, 30, main="df=5")
hist(x3, 30, main="df=15")
```

We see that: - Small df implies that values are more concentrated around
0, with a long tail - Large df implies that values are more spread out,
with a shorter tail

### Poisson distribution: change lambda

```{r}
x1 <- rpois(n, lambda=1)
x2 <- rpois(n, lambda=5)
x3 <- rpois(n, lambda=15)

par(mfrow=c(3,1))

hist(x1, 30, main="lambda=1")
hist(x2, 30, main="lambda=5")
hist(x3, 30, main="lambda=15")
```

We see that: - Small lambda implies that values are more concentrated
around 0, with a long tail - Large lambda implies that values are more
spread out, with a shorter tail

### t distribution: change the degrees of freedom

```{r}
x1 <- rt(n, df=1)
x2 <- rt(n, df=5)
x3 <- rt(n, df=30)

par(mfrow=c(3,1))

hist(x1, 30, main="df=1")
hist(x2, 30, main="df=5")
hist(x3, 30, main="df=30")
```

We see that: - Small df implies that values are more concentrated around
0, with a long tail - Large df implies that values are more spread out,
with a shorter tail, and when df is large the distribution looks more
like a normal distribution

Conclusion: moving the parameters change - location -\> mean - spread
-\> variance/sd - skewness -\> rate/df - tail -\> rate/df

Observe what happens when you move the parameters. For different
distributions, do for example

```{r}
n <- 1000
k <- 30
x <- rexp(n, rate=2)
plot(x, type="l")
hist(x, k)
```

Here k denotes the number of classes for the histogram. Test with other
distributions writing distributions in the help utility.

To explore all distributions in R:

```{r}
help("Distributions")
```

or

```{r}
?rexp
?rnorm
```

## 2. Verify the IID character of any sample of the previous exercise. Use Ljung-Box test with the instruction

```{r}
h <- 1
Box.test (x, lag=h, type="Ljung")
```

For example:

```{r}
n <- 10000
k <- 3
x <- rt(n, df=k)
plot(x, type="l")
Box.test(x, lag=1, type="Ljung")
```

Interpretation: We see a large p-value, which means that we cannot
reject the null hypothesis of independence, so we can conclude that the
data are independent. The sample behaves like IID noise, as expected.

We can try different lags, from h=1, to log(n):

```{r}
for(h in 1:log(n)){
  print(Box.test(x, lag=h, type="Ljung"))
}
```

We see that for all lags, we cannot reject the null hypothesis of
independence, so we can conclude that the data are independent. p-values
are all large, confirming that the data are independent.

### Comparison with random walk

A random walk is not IID because each value depends on the previous one.
We can verify this by applying the Ljung-Box test to a random walk:

```{r}
n <- 10000
rw <- cumsum(rnorm(n))
plot(rw, type="l")
Box.test(rw, lag=log(n), type="Ljung")
```

We see a small p-value, which means that we reject the null hypothesis
of independence, so we can conclude that the data are not independent.
The random walk does not behave like IID noise, as expected.

We can try different lags for the random walk as well:

```{r}
rw <- cumsum(rnorm(n))
for(h in 1:log(n)){
  print(Box.test(rw, lag=h, type="Ljung"))
}
```

We see that for all lags, we reject the null hypothesis of independence,
so we can conclude that the data are not independent. p-values are all
small, confirming that the data are not independent.

# Practical 2B

## 1. Generate an IID noise of length n and find its exponential smoothing series

The following piece of code generates the Holt Winters smoothing for
stationary series with no trend and no seasonality.

```{r}
n <- 100
x <- rnorm(n)
plot(x,
     main = "IID noise of length 100",
     sub = "n = 100",
     xlab = "Time",
     ylab = "Noise")

a <- 0.1
y <- HoltWinters(x,alpha=a,beta=F,gamma=F)
plot(y, main = "Holt-Winters filtering (a = 0.1)")
```

The generated series is an IID noise, so it does not have any systematic
trend or seasonal component.

When generating the HoltWinters filtering, we set beta=FALSE and
gamma=FALSE, which implies that the Holt-Winters method simply applies
an exponential smoothing, which estimates a smoothed level of the
series. In fact, the filtered signal, which is **red**, oscillates
around the mean of the noise and removes part of the high-frequency
variability. Next we will see how this behavior of removing the
high-frequency components changes as we vary "a".

### 1.1 Try for different values of: "n" and “a”

First we create a function to plot 6 subplots in one view, so that we
can easily visualize how the filter changes when "a" varies:

```{r}
hw_noise_plots <- function(n){

  set.seed(123) # we fix the seed so that we have the same noise each time we run the code
  
  # generate IID noise
  x <- rnorm(n)

  # alpha values
  alphas <- c(0.01, 0.1, 0.2, 0.5, 0.8, 0.9)

  # layout
  par(
  mfrow = c(2,3),
  oma = c(0,0,3,0),   # outer margin for global title
  mar = c(4,4,2,1)    # bottom, left, top, right margins
  )

  # loop over alphas
  for(a in alphas){

    y <- HoltWinters(x, alpha = a, beta = FALSE, gamma = FALSE)

    plot(y, main = paste("a =", a))
  }

  # global title
  mtext(paste("Holt-Winters filtering with n =", n),
        outer = TRUE, cex = 1.3)
}
```

For n high (e.g. n\>200) it's hard to see the filter signal on top of
the original time series because the time series in very dense, so its
black color prevails over the red. For this reason, we create another
function to plot the original series in **grey**, and add on top the
filtered signal in **red**:

```{r}
hw_noise_plots_n_high <- function(n,x){

  set.seed(123)
  x <- rnorm(n)

  alphas <- c(0.01, 0.1, 0.2, 0.5, 0.8, 0.9)

  par(
  mfrow = c(2,3),
  oma = c(0,0,3,0),   # outer margin for global title
  mar = c(4,4,2,1)    # bottom, left, top, right margins
  )

  for(a in alphas){

    y <- HoltWinters(x, alpha=a, beta=FALSE, gamma=FALSE)
    ylim <- range(x) # to fix same y-scale in the plot

    # plot noise
    plot(x,
         ylim = ylim,
         type="l",
         col="grey",
         main=paste("alpha =", a),
         xlab="Time",
         ylab="Value")

    # add smoothed signal
    lines(y$fitted[,1], col="red", lwd=2)
  }

  mtext(paste("Holt-Winters filtering with n =", n),
        outer=TRUE, cex=1.3)
}
```

Now we will try to vary "a" between 0 and 1, for different amount of
data points n (from 10 to 1000). The aim is to compare how the filtering
behavior changes with "a" small or large:

```{r}
# n = 10, a varies between 0 and 1
hw_noise_plots(10)

# n = 50, a varies between 0 and 1
hw_noise_plots(50)

# n = 100, a varies between 0 and 1
hw_noise_plots(100) 

# n = 200, a varies between 0 and 1
hw_noise_plots_n_high(200) # I use the function to plot the original series in grey so that I can clearly see the red filtered signal on top of it

# n = 300, a varies between 0 and 1
hw_noise_plots_n_high(300) 

# n = 500, a varies between 0 and 1
hw_noise_plots_n_high(500) 

# n = 1000, a varies between 0 and 1
hw_noise_plots_n_high(1000) 
```

**Observations:**

***Changing "a":*** As it is confirmed by the plots above, the parameter
"a" controls how smooth is the estimation of the series:

-   Small values of "a" produce a very smooth estimate

-   Larger values of "a" produce a very noisy estimate, that tends to
    match the original series

This is coherent with the fact that a small "a" gives more weight to
past observations, while a large "a" gives more weight to recent
observations and so the filtered series reacts quickly to changes in
noise - and therefore for "a" very large the filtered series almost
matches the original series.

***Changing n:*** this doesn't change the statistical properties of the
IID noise itself, but it's more a matter of visualization and
observation:

-   for n small, we can see the filtered series much more clearly
    because the plotted signal is not dense; so it's easier to visualize
    how far is the filtered signal from the original series. But on the
    other side, we can observe the behavior of the smoothing filter for
    only few data points (e.g. 50);

-   for n large, the plotted signal is visualized as much more dense so
    it's not easy to see all the peaks and variations of the noise in
    detail, especially when "a" increases as well. So it may be harder
    to visualize the effect of the smoothing filter, but on the other
    hand it allows to observe the behavior of the filter over longer
    periods of time (i.e. for more data points, e.g. 500, 1000).

### 1.2 Simulate other IID noises with other laws. If do not prefix “a” the parameter is estimated.

The exponential smoothing can also be applied to IID noises generated
from other distributions. In the following, we propose to apply it to
noises generated with **uniform, exponential, Poisson, and Student-t
distributions**.

As we commented above, changing n doesn't affect the statistical
properties of the IID noise, so below we will plot the estimated filter
for different values of "a" while we keep n = 100.

The "a" values we will plot are: 0.01, 0.1, 0.2, 0.5, 0.8, 0.9.

-   **Uniform noise**

```{r}

# Uniform noise between -1 and 1

# First define the function to plot for varying values of "a"
hw_noise_plots_uniform <- function(n){

  set.seed(123)
  x <- runif(n, min = -1, max = 1)

  alphas <- c(0.01, 0.1, 0.2, 0.5, 0.8, 0.9)

  par(
  mfrow = c(2,3),
  oma = c(0,0,3,0),   # outer margin for global title
  mar = c(4,4,2,1)    # bottom, left, top, right margins
  )

  for(a in alphas){

    y <- HoltWinters(x, alpha=a, beta=FALSE, gamma=FALSE)
    ylim <- range(x) # to fix same y-scale in the plot
    # plot noise
    plot(x,
         type="l",
         ylim = ylim,
         col="grey",
         main=paste("alpha =", a),
         xlab="Time",
         ylab="Value")

    # add smoothed signal
    lines(y$fitted[,1], col="red", lwd=2)
  }

  mtext(paste("Holt-Winters Uniform Noise filtering with n =", n),
        outer=TRUE, cex=1.3)
}  

```

```{r}
hw_noise_plots_uniform(100) 

```

-   **Exponential noise (positive skewed noise)**

```{r}

# Exponential noise (positive skewed noise)

# First define the function to plot for varying values of "a"
hw_noise_plots_exp <- function(n){

  set.seed(123)
  x <- rexp(n, rate = 1)

  alphas <- c(0.01, 0.1, 0.2, 0.5, 0.8, 0.9)

  par(
  mfrow = c(2,3),
  oma = c(0,0,3,0),   # outer margin for global title
  mar = c(4,4,2,1)    # bottom, left, top, right margins
  )

  for(a in alphas){

    y <- HoltWinters(x, alpha=a, beta=FALSE, gamma=FALSE)
    ylim <- range(x) # to fix same y-scale in the plot
    # plot noise
    plot(x,
         ylim = ylim,
         type="l",
         col="grey",
         main=paste("alpha =", a),
         xlab="Time",
         ylab="Value")

    # add smoothed signal
    lines(y$fitted[,1], col="red", lwd=2)
  }

  mtext(paste("Holt-Winters Exponential Noise filtering with n =", n),
        outer=TRUE, cex=1.3)
}  

hw_noise_plots_exp(100)
```

-   **Poisson noise (discrete IID noise)**

```{r}

# Poisson noise

# First define the function to plot for varying values of "a"
hw_noise_plots_pois <- function(n){

  set.seed(123)
  x <- rpois(n, lambda = 3)

  alphas <- c(0.01, 0.1, 0.2, 0.5, 0.8, 0.9)

  par(
  mfrow = c(2,3),
  oma = c(0,0,3,0),   # outer margin for global title
  mar = c(4,4,2,1)    # bottom, left, top, right margins
  )

  for(a in alphas){

    y <- HoltWinters(x, alpha=a, beta=FALSE, gamma=FALSE)
    ylim <- range(x) # to fix same y-scale in the plot
    
    # plot noise
    plot(x,
         ylim = ylim,
         type="l",
         col="grey",
         main=paste("alpha =", a),
         xlab="Time",
         ylab="Value")
    
    # add smoothed signal
    lines(y$fitted[,1], col="red", lwd=2)
  }

  mtext(paste("Holt-Winters Poisson Noise filtering with n =", n),
        outer=TRUE, cex=1.3)
}  

hw_noise_plots_pois(100)
```

-   **Student-t noise**

```{r}

# Student-t noise

# First define the function to plot for varying values of "a"
hw_noise_plots_stud <- function(n){

  set.seed(123)
  x <- rt(n, df = 5)

  alphas <- c(0.01, 0.1, 0.2, 0.5, 0.8, 0.9)

  par(
  mfrow = c(2,3),
  oma = c(0,0,3,0),   # outer margin for global title
  mar = c(4,4,2,1)    # bottom, left, top, right margins
  )

  for(a in alphas){

    y <- HoltWinters(x, alpha=a, beta=FALSE, gamma=FALSE)
    ylim <- range(x) # to fix same y-scale in the plot
    
    # plot noise
    plot(x,
         ylim = ylim,
         type="l",
         col="grey",
         main=paste("alpha =", a),
         xlab="Time",
         ylab="Value")
    
    # add smoothed signal
    lines(y$fitted[,1], col="red", lwd=2)
  }

  mtext(paste("Holt-Winters Student-t Noise filtering with n =", n),
        outer=TRUE, cex=1.3)
}  

hw_noise_plots_stud(100)
```

**Observations:**

The general behavior of the filter remains similar:

-   Small "a" (e.g. 0.01, 0.1) –\> more weight to past observations,
    very smooth estimate

-   Large "a" (e.g. 0.8, 0.9) –\> more weight to recent observations, so
    the filtered series reacts quickly to changes in noise –\> noisy
    estimate

What changes with respect to **x \<- rnorm(n)** is that the shape of the
smoothed signal reflects the characteristics of the original
distribution.

## 2. Apply Holt Winters smoothing to real data with trend and seasonality

```{r}
x<-AirPassengers
a<-0.2
b<-0.3
c<-0.8
y<-HoltWinters(x,alpha=a,beta=b, gamma=c)
plot(y, main = "Holt-Winters filtering (AirPassengers dataset)")
```

The AirPassengers dataset shows a strong upward trend and a clear yearly
seasonality.

The Holt-Winters method is therefore appropriate here, because it
internally decomposes the series into level, trend, and seasonal
components. The fitted values (red line) follow the upward trend over
time and replicate the seasonal pattern, while applying a soft smoothing
effect. In fact, the red curve fluctuates less than the original black
series.

We will now see the role that the parameters "a", "b", "c" play in
determining how quickly the model updates the level, trend, and seasonal
components:

-   "a" determines how quickly the **level** adapts to new
    observations.\
    With **"a = 0.2"** the level adapts gradually to new observations -
    while still giving substantial weight to the past ones.

-   "b" determines how quickly the **trend** component adapts.\
    With **"b = 0.3"** the trend evolves smoothly over time while still
    responding to changes in the trend of the series. This results in a
    trend that is smoother than the original signal but still closely
    aligned with it.

-   "c" determines how quickly the **seasonal** component is updated.\
    With **"c = 0.8"**, high weight is given to the recent seasonal
    observations, therefore the seasonal pattern adapts quickly to
    changes in seasonal amplitude.

As a result, the fitted curve closely follows the repeating yearly
seasonal pattern while smoothing the fluctuations.

### 2.1 Change the series and use for example log(x), diff(log(x)), diff(x,12), etc.

The exercise suggests to now try to use the HoltWinters filter on
transformations of the original signals, such as log(x), diff(log(x)),
diff(x,12).

The transformations of the series can be applied in order to stabilize
variance or remove trend and seasonality. For example:

-   The logarithmic transformation reduces the growth of seasonal
    amplitude over time

-   The first difference removes part of the trend

-   The seasonal difference with lag 12 removes the yearly seasonal
    pattern

We will also use the HoltWinters on additional transformations:

-   **diff(log(AirPassengers), 12)** to combine the effect of the
    logarithmic transformation and seasonal difference with lag 12.

-   **diff(diff(log(AirPassengers), 12)** to combine the effect of the
    first difference, with logarithmic transformation and seasonal
    difference with lag 12.

```{r}
x0<-AirPassengers
a<-0.2
b<-0.3
c<-0.8

x1 <- log(x0)
x2 <- diff(log(x0))
x3 <- diff(x0, 12)
x4 <- diff(log(x0), 12)
x5 <- diff(diff(log(x0), 12))

# Log(x)
y<-HoltWinters(x1,alpha=a,beta=b, gamma=c)
plot(y, main = "Holt-Winters filtering with log(AirPassengers)")

# For the differencing transformations, the trend and seasonality are already removed, so we will set beta and gamma FALSE

# diff(log(x))
y<-HoltWinters(x2,alpha=a,beta=F, gamma=F)
plot(y, main = "Holt-Winters filtering with diff(log(AirPassengers))")

# diff(x,12)
y<-HoltWinters(x3,alpha=a,beta=F, gamma=F)
plot(y, main = "Holt-Winters filtering with diff(AirPassengers,12)")

# diff(log(AirPassengers), 12)
y<-HoltWinters(x4,alpha=a,beta=F, gamma=F)
plot(y, main = "Holt-Winters filtering with diff(log(AirPassengers),12)")

# diff(diff(log(AirPassengers), 12)
y<-HoltWinters(x5,alpha=a,beta=F, gamma=F)
plot(y, main = "Holt-Winters filtering with diff(diff(log(AirPassengers), 12))")
```

**Observations:**

1.  **log(AirPassengers)\
    **While in the original series the seasonal peaks becomes larger
    over time (initial oscillations between 100 and 200, final
    oscillations between 400 and 600), in the log-transformed series the
    oscillations are more stable (initial oscillations between 0 and 5,
    final oscillations between 6 and 6.5 approximately).\
    As a consequence, the Holt-Winters fitted curve therefore follows
    the trend and seasonal pattern more uniformly across the whole time
    range.\

2.  **diff(log(AirPassengers))\
    **After applying the first difference to the log-transformed series,
    the trend component is removed. The series now oscillates around 0,
    and the Holt-Winters filter captures more the signal variations
    rather than the upward trend.\

3.  **diff(AirPassengers, 12)\
    **The seasonal difference with lag 12 tends to remove the yearly
    seasonal component. In fact, the resulting series no longer shows
    the repeating yearly peaks that were visible in the original data.
    The filtered series therefore highlights the remaining
    fluctuations.\

4.  **diff(log(AirPassengers), 12)**

    Here we combined the logarithmic transformation with the seasonal
    difference: so we stabilized the variance and removed the yearly
    seasonal pattern at the same time.\
    The resulting series appears more stationary, fluctuating between 0
    and 0.3 approximately.\
    The Holt-Winters filter therefore produces a smoother signal that
    fluctuates around 0.1, capturing the remaining variations in the
    series.\

5.  **diff(diff(log(AirPassengers), 12))**

    In this case we applied an additional first difference to the
    transformation of point 4. This further removes any remaining trend
    component. So the idea was to see if in the signal of point 4 there
    was any remaining trend component.\
    The plots of point 4 and 5 indeed show a different resulting series:
    the first difference produced a more stationary signal, with
    fluctuations centered around zero. \
    In this case the Holt-Winters filter smoothens the fluctuations
    rather than modeling trend or seasonality, and it captures the few
    remaining variations in the series.

### 2.2 Change the values of parameters a, b and c

For every combination of a,b and c, we will plot the Holt-Winters
filtering on the above transformations 1, 2, 3, 4, 5, 6.

To avoid repeating the code every time, we first create a function that
takes as input a, b and c. We will plot in the first line the
AirPassengers and Log(AirPassengers) with the same scale; and in the
other lines the other 4 transformations with a same scale between them.
In this way we can easily see the changes between transformations. We
won't keep the same scale between all the 6 plots because AirPassengers
and Log(AirPassengers) have much larger scale, so the other 4 plots
would be visible as lines (all information of the visualization would be
lost).

```{r}
plot_hw_transforms <- function(a, b, c) {
  
  # Series
  x1 <- AirPassengers
  x2 <- log(x1)
  x3 <- diff(log(x1))
  x4 <- diff(x1, 12)
  x5 <- diff(log(x1), 12)
  x6 <- diff(diff(log(x1), 12))
  
  # Holt-Winters fits
  y1 <- HoltWinters(x1, alpha = a, beta = b, gamma = c)
  y2 <- HoltWinters(x2, alpha = a, beta = b, gamma = c)
  y3 <- HoltWinters(x3, alpha = a, beta = b, gamma = c)
  y4 <- HoltWinters(x4, alpha = a, beta = b, gamma = c)
  y5 <- HoltWinters(x5, alpha = a, beta = b, gamma = c)
  y6 <- HoltWinters(x6, alpha = a, beta = b, gamma = c)
  
  # Shared scale for large-magnitude series
  ylim_big <- range(
    c(x1, x4, y1$fitted[,1], y4$fitted[,1]),
    na.rm = TRUE
  )
  
  # Shared scale for log-based transformed series
  ylim_small <- range(
    c(x2, x3, x5, x6,
      y2$fitted[,1], y3$fitted[,1], y5$fitted[,1], y6$fitted[,1]),
    na.rm = TRUE
  )
  
  # Layout: 2 plots on top, 4 below
  layout(matrix(c(1,2,
                  3,4,
                  5,6), nrow = 3, byrow = TRUE))
  
  par(oma = c(0, 0, 3, 0), mar = c(4, 4, 3, 1))
  
  plot_hw <- function(x, y, title, ylim) {
    plot(x, col = "grey60", main = title, ylim = ylim)
    lines(y$fitted[,1], col = "red", lwd = 2)
  }
  
  # Top row
  plot_hw(x1, y1, "AirPassengers", ylim_big)
  plot_hw(x4, y4, "diff(AirPassengers, 12)", ylim_big)
  
  # Bottom 4
  plot_hw(x2, y2, "log(AirPassengers)", ylim_small)
  plot_hw(x3, y3, "diff(log(AirPassengers))", ylim_small)
  plot_hw(x5, y5, "diff(log(AirPassengers), 12)", ylim_small)
  plot_hw(x6, y6, "diff(diff(log(AirPassengers), 12))", ylim_small)
  
  mtext(paste("a =", a, ", b =", b, ", c =", c), outer = TRUE, cex = 1.3)
}


```

```{r}

# Initial values of a, b, c
plot_hw_transforms(0.2, 0.3, 0.8)

# Decrease only a (a was 0.2)
plot_hw_transforms(0.05, 0.3, 0.8)

# Decrease only b (b was 0.3)
plot_hw_transforms(0.2, 0.1, 0.8)

# Decrease only c (c was 0.8)
plot_hw_transforms(0.2, 0.3, 0.4)

# Decrease a, b, c
plot_hw_transforms(0.05, 0.1, 0.4)

# Increase only a
plot_hw_transforms(0.5, 0.3, 0.8)

# Increase only b
plot_hw_transforms(0.2, 0.6, 0.8)

# Increase only c
plot_hw_transforms(0.2, 0.3, 0.9)

# Increase a, b, c
plot_hw_transforms(0.5, 0.6, 0.9)

```

Since the original series and the transformed series have very different
magnitudes, using a single common scale for all subplots makes the
transformed signals hard to interpret. Therefore, we decided to use a
common scale between AirPassengers and Log(AirPassengers) and a common
scale between the other transformations.

As we saw before, "a" controls the smoothing of the level, "b" of the
trend, "c" of the seasonality.

So in the plots above we tried to first decrease/increase each parameter
individually, and then all of them together:

-   **Decrease (increase) only a –\> plot 2 (plot 6)**\
    The level reacts slowly (quickly) to new observations. The fitted
    curve follows the data more gradually (closely) and adapts slowly
    (rapidly) to changes in the series.

-   **Decrease (increase) only b –\> plot 3 (plot 7)**\
    The trend component changes more slowly (faster) over time: the
    estimated trend becomes smoother and less sensitive to variations in
    the data (adjusts faster to changes in the slope of the series).

-   **Decrease (increase) only c –\> plot 4 (plot 8)**\
    The seasonal component becomes more stable (adapts more quickly to
    recent seasonal patterns). The seasonal pattern changes less over
    time (becomes more reactive to new observations) and the model gives
    more weight to past (recent) seasonal estimates.

<!-- -->

-   **Decrease (increase) a, b, c –\> plot 5 (plot 9)**\
    When all parameters decrease (increase), the model reacts slowly
    (faster) to new information. The fitted curve adapts gradually
    (quickly) to changes in level, trend, and seasonality.

## 3. Holt Winters prediction

Now we will see that the Holt-Winters method can also be used to
forecast future observations.

```{r}
x<-AirPassengers
a<-0.3
b<-0.1
c<-0.8
y<-HoltWinters(x,alpha=a,beta=b,gamma=c)
plot(y, main = "Holt Winters filtering for AirPassengers")

z<-predict(y,n.ahead=4*12)
plot(y,z,lty=1:2, main = "Holt Winters filtering and prediction for AirPassengers")

```

As we can see in the plot above, the red line represents:

-   on the left of the dashed line –\> filter on the observed time
    series

-   on the right of the dashed line –\> prediction for the next 4 years
    ( predict(y,n.ahead=4\*12) )

The forecast shows a continuation of the upward trend together with the
repeating yearly seasonal pattern. Also, the amplitude of the seasonal
fluctuations keeps increases as the level of the series grows.

## 4. "a", "b", "c" are not prefixed

In this case we can create the Holt-Winters filter without giving
explicit values of a, b and c. They will be automatically estimated, so
we will use the commands "y\$alpha", "y\$beta", "y\$gamma" to read the
estimated values.

```{r}
x<-AirPassengers
y<-HoltWinters(x)
y$alpha
y$beta
y$gamma
plot(y)
z<-predict(y,n.ahead=4*12)
plot(y,z,lty=1:2)
```

In order to minimize the forecast error, and to provide an optimal
balance between responsiveness to new observations and stability of the
level, trend, and seasonal components, the Holt-Winters model estimated
the optimal parameters to be:

```         
- alpha = 0.2479595 
- beta = 0.03453373 
- gamma = 1 
```

# 5. Wine sales in Australia

We will now see the effect of the Holt-Winters filtering and prediction
on different data sets.

```{r}
x<-wineind
y<-HoltWinters(x)
plot(x, main = "wineind dataset")
y$alpha
y$beta
y$gamma
plot(y)
z<-predict(y,n.ahead=4*12)
plot(y,z,lty=1:2, main = "Holt-Winters filtering and prediction")
```

**Observations:**

In this case, the Holt-Winters method automatically estimates the
parameters as:

```         
- alpha = 0.01655901 
- beta = 0.812572 
- gamma = 0.2759254 
```

**Automatic parameters selection:**\
"a" is very small, meaning that the level component changes slowly and
gives strong weight to past observations. "b" is large (close to 1),
which makes the model adapt quickly to changes in the trend. "c" takes a
moderate value (around 0.27), meaning that the seasonal pattern is
updated gradually over time rather than reacting too strongly to the
fluctuations.\

**Forecast**: \
The forecast continues the seasonal fluctuations with approximately the
same amplitude, indicating that the model expects the seasonal structure
of the series to remain stable in the future.

### 5.1 other examples

We will now try Holt-Winters on more data sets:

-   Monthly atmospheric CO₂ concentration
-   UK gas consumption
-   Monthly temperature - Nottingham

```{r}
# Monthly atmospheric CO₂ concentration

x <- co2
y <- HoltWinters(x)

plot(x, main = "Monthly atmospheric CO2 concentration")
y$alpha
y$beta
y$gamma
plot(y)
z<-predict(y,n.ahead=4*12)
plot(y,z,lty=1:2, main = "Holt-Winters filtering and prediction")
```

The parameters "a", "b", "c" automatically selected create a fitting
curve that describes well the original series (the red line maintains
the level, trend and seasonality). As a consequence, the forecast
produced for the next four years continues the same structural pattern,
showing a consistent evolution of the level together with the repeating
seasonal oscillations.

-   **UK gas consumption**

```{r}
# UK gas consumption

x <- UKgas
y <- HoltWinters(x)

plot(x, main = "UK gas consumption")
y$alpha
y$beta
y$gamma
plot(y)
z<-predict(y,n.ahead=4*12)
plot(y,z,lty=1:2, main = "Holt-Winters filtering and prediction")
```

The series of *UK gas consumption* shows a strong increasing trend
together with a seasonal pattern whose amplitude grows over time. The
forecast continues the same structural behaviour: the predicted values
show a continuation of the increasing trend together with seasonal
oscillations whose amplitude keeps growing over time.

-   **Monthly temperature - Nottingham**

```{r}
# Monthly temperature - Nottingham
x <- nottem
y <- HoltWinters(x)

plot(x, main = "Monthly temperature - Nottingham")
y$alpha
y$beta
y$gamma
plot(y)
z<-predict(y,n.ahead=4*12)
plot(y,z,lty=1:2, main = "Holt-Winters filtering and prediction")
```

The *Monthly temperature - Nottingham* series shows a clear seasonal
pattern but not a long-term trend (the fluctuations are around a
relatively stable level). The Holt–Winters filtering captures this
behavior well, and in fact the fitted curve follows the seasonal pattern
while smoothing part of the irregular fluctuations in the original
observations.
