---
title: "Practical 3 - AR(p) Models"
authors: Eleonora Mazzocca, Nùria Torquet
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

# Practical 3a AR(p) Models

## 1. Simulating AR(p) models

Generate an AR(2) model:

```{r}

n<-1000
x<-arima.sim(list(ar = c(0.8897, -0.4858), sd = sqrt(0.1796)),n)
plot(x,type="l",main="Simulated AR(2) Time Series", ylab="Value", xlab="Time")

```

The simulated series shows a stochastic behavior with no clear trend or
seasonality, as expected from an AR(p) process. The values fluctuate
around a constant mean, suggesting stationarity.

Compute its autocorrelation function:

```{r}
acf(x, main="ACF of Simulated AR(2)")
```

The autocorrelation function shows a gradual decay (kind of
exponential), which is characteristic of AR processes. This confirms
that the data was generated from an AR model.

### Repeat for different n, number of parameters and different p of the AR(p).

First I create a function to build the AR(p) model, so that I can easily
vary n, p and the parameters phi:

```{r}
simulate_ar <- function(n, phi, sigma2 = 1) {
  
  x <- arima.sim(
    model = list(ar = phi),
    n = n,
    sd = sqrt(sigma2)
  )
  
  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par))
  
  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  
  plot(
    x,
    type = "l",
    main = paste0("AR(", length(phi), "), n = ", n),
    xlab = "Time",
    ylab = "Value"
  )
  
  acf(x, main = "ACF")
}
```

### REPEAT FOR DIFFERENT n

```{r}
n_values <- c(100, 500, 1000)

for (n in n_values) {
  simulate_ar(n, phi = c(0.8897, -0.4858), sigma2 = 0.1796)
  dev.flush()
}
```

*Observations:*

-   Smaller n (100): the simulated series appears more irregular and the
    ACF is noisier. This makes the underlying auto-regressive structure
    less clear, because of the visual impact of the random fluctuations.

-   Larger n (e.g. 1000): the behavior of the series becomes more
    representative of the theoretical AR model. The expected decay
    pattern of the ACF becomes clearer and smoother (after the decay,
    the values tend to be around 0, while for n=100 the values were
    still oscillating between -0.2 and 0.2 --\> dashed blue lines).

Overall, increasing n does not change the true model, but it improves
the visibility of its properties.

### REPEAT FOR DIFFERENT AR orders

```{r}
phi_list <- list(
  c(0.7), #AR(1) 
  c(0.8897, -0.4858), #AR(2)
  c(0.6, -0.3, 0.2) #AR(3)
)

for (phi in phi_list) {
  simulate_ar(1000, phi)
}
```

Changing the order *p* changes how many past observations influence the
current value. As p increases, the dependence structure becomes more
complex and may persist over more lags.

-   **AR(1):** each observation depends only on the immediately previous
    value. The ACF has a smooth decay.

-   **AR(2)** and **AR(3):** the dependence involves more than one lag,
    so the series show richer dynamics. The ACF still decays gradually,
    but less smoothly than the AR(1) case and with an oscillatory
    behavior (resembling damped oscillations).

Overall, increasing the order of the model *p* increases the complexity
of temporal dependence. This can be observed **both in the time series
plot** (signals are reacher as *p* increases) **and in the ACF**, which
shows less smooth decay and more complex shapes due to the combined
effect of multiple lagged terms.

### REPEAT FOR DIFFERENT p

I choose to fix n = 1000 and p=2, so that I can observe how the time
series and ACF change as I vary the parameters.

```{r}
set.seed(123)
n <- 500

phi_list_values <- list(
  c(0.3, 0.2),
  c(0.8897, -0.4858),
  c(1.2, -0.7)
)

titles_values <- c(
  "AR(2): phi = (0.3, 0.2)",
  "AR(2): phi = (0.8897, -0.4858)",
  "AR(2): phi = (1.2, -0.7)"
)

par(mfrow = c(3, 2), mar = c(4, 4, 3, 1))

for (i in 1:length(phi_list_values)) {
  
  x <- arima.sim(list(ar = phi_list_values[[i]]), n = n, sd = sqrt(0.1796))
  
  plot(x, type = "l",
       main = paste(titles_values[i], "- Series"),
       xlab = "Time", ylab = "Value")
  
  acf(x,
      main = paste(titles_values[i], "- ACF"))
}

par(mfrow = c(1, 1))
```

**Observations:**

The three simulated AR(2) series show:

-   similar overall behavior –\> in fact we can observe stationary
    fluctuations around a constant mean

-   their dependence structure changes depending on the parameter values

    -   phi = (0.3, 0.2) –\> the series looks "random" and the ACF goes
        to zero smoothly.\
        This is the case with small coefficients, so the dependence
        between observations is weak.

    -   phi = (0.8897, -0.4858) –\> the series looks slightly smoother
        than previous, the decay is a bit more oscillatory. The ACF
        stabilizes around 0 in longer time than phi = (0.3, 0.2) - which
        is in accordance with larger coefficients leading to stronger
        persistence.

    -   phi = (1.2, -0.7) –\> the series looks more structured and less
        random, and the ACF has stronger oscillations. This could be
        indicative of alternating correlations across lags - which is
        compatible with such large coefficients.

Overall, increasing the magnitude of the coefficients increases the
persistence of the process. This also leads to oscillatory behavior in
the ACF, which indicates a more complex temporal dependence.

## 2. Fitting AR(p) models

With the following commands, we will obtain an estimation of the AR(2)
parameters and the intercept. We will assess the quality of the fit, and
try with larger values of n.

```{r}
n<-1000
x<-arima.sim(list(ar = c(0.8897, -0.4858), sd = sqrt(0.1796)), n)
plot(x,type="l",main="Simulated AR(2) series",xlab="Time",ylab="Value")
s1<-arima(x, order=c(2,0,0), method="ML")
s1
```

**Observations:**

-   As it happened before, the generated time series fluctuates around a
    constant mean with no visible trend or seasonality, which is
    consistent with a stationary AR(2) process.

-   **Fitted values:** these values estimate the autoregressive
    coefficients from the simulated data and they resulted to be very
    close to the true values

    -   true coefficients: ϕ1=0.8897 and ϕ2=−0.4858

    -   estimated coefficients: ϕ1=0.8665 and ϕ2=−0.4481

        So the model fits the data well.

### Try with larger values of n

```{r}
n_values <- c(100, 500, 1000, 5000)

par(mfrow = c(2,2), mar = c(4,4,3,1))

for (n in n_values) {
  x <- arima.sim(list(ar = c(0.8897, -0.4858), sd = sqrt(0.1796)), n = n)
  
  plot(x, type = "l",
       main = paste("Simulated AR(2), n =", n),
       xlab = "Time",
       ylab = "Value")
  
  s1 <- arima(x, order = c(2,0,0), method = "ML")
  
  cat("\n========================\n")
  cat("n =", n, "\n")
  print(s1)
}

par(mfrow = c(1,1))
```

**Observations:**

From the results above, we can notice that changing *n* brings to really
different performance of the estimating model. As the number of samples
*n* increases, the estimated AR coefficients gets closer to the real
coefficients. In fact, the standard error (s.e.) for ar1, ar2 and
intercept decreases. Moreover, as *n* increases, also the performance
indicator *aic* increases. So we can conclude that if the model has more
data points, it will be easier to be more accurate in estimating the
autoregressive coefficients.

### Do it several times simulating new series with different parameters

```{r}

n <- 1000

phi_list <- list(
  c(0.5, 0.2),
  c(0.8897, -0.4858),
  c(1.2, -0.7)
)

par(mfrow = c(3,1), mar = c(4,4,3,1))

for (phi in phi_list) {
  x <- arima.sim(list(ar = phi, sd = sqrt(0.1796)), n = n)
  
  plot(x, type = "l",
       main = paste("Simulated AR(2): phi = (", paste(phi, collapse = ", "), ")", sep = ""),
       xlab = "Time",
       ylab = "Value")
}

par(mfrow = c(1,1))

for (phi in phi_list) {
  x <- arima.sim(list(ar = phi, sd = sqrt(0.1796)), n = n)
  s1 <- arima(x, order = c(2,0,0), method = "ML")
  
  cat("\n========================\n")
  cat("True phi =", paste(phi, collapse = ", "), "\n")
  print(s1)
}
```

**Observations:**

In this case as well, with different simulated time series, the fitting
model performs well and estimates the autoregression coefficients which
are very close to the true coefficients (all the standard errors on ar1,
ar2, intercept are very small (around 0.01-0.06).

This shows that the fitted model adapts to different structures: in all
these cases it is able to recover the main autoregressive structure.

### Try alternative method

Try instead s2\<- ar(x, order.max=2, method=”mle”).

```{r}
n <- 1000
x <- arima.sim(list(ar = c(0.8897, -0.4858), sd = sqrt(0.1796)), n = n)

s1 <- arima(x, order = c(2,0,0), method = "ML")
s2 <- ar(x, order.max = 2, method = "mle")

s1
s2
```

**Observations:**

The two method provide the same autoregressive coefficients and
sigma\^2.

The first method (s1) might be more helpful as it shows explicitly more
details, e.g. the standard error and the intercept.

### Try order.max=3

```{r}
s3 <- ar(x, order.max = 3, method = "mle")
s3
```

In this case the fitting model tries AR up to the 3rd order: AR(1),
AR(2) and AR(3) and compares them using *aic* internally, to then choose
the best one.

As the result shows, the best model chosen is still AR(2) and so it
returned the same coefficients as before.

This makes sense because the time series that we are fitting, x, was
generated from AR(2). So the model doesn't need to represent it with
AR(3), indicating that adding an extra lag does not improve the model
fit.

Overall, even when a higher maximum order is allowed, the model
selection procedure correctly identifies AR(2) as the best model.

## 3. Validation

We will now experiment with the Box.test functionality.

First we create an AR(2) model with phi = (0.8897, -0.4858).

```{r}
n<-1000
x<-arima.sim(list(ar = c(0.8897, -0.4858), sd = sqrt(0.1796)), n)
plot(x,type="l",main="Simulated AR(2) series, phi = (0.8897, -0.4858)",xlab="Time",ylab="Value")
s<-arima(x, order=c(2,0,0), method="ML")
s
```

```{r}
y<-resid(s)

# Residuals plot + ACF together
par(mfrow = c(3,1))
plot(x,type="l",main="Simulated AR(2) series, phi = (0.8897, -0.4858)",xlab="Time",ylab="Value")
plot(y,type="l",main = "Residuals", xlab = "Time", ylab = "Residuals")
acf(y, main = "ACF of Residuals")
par(mfrow = c(1,1))

# Ljung-Box tests
h<-floor(log(n))
Box.test(x,lag=h, type=c("Ljung-Box"))
Box.test(y,lag=h,type=c("Ljung-Box"))
```

**Observations:**

The residuals (y) fluctuate randomly around zero with no visible trend
or structure, which suggests that the model has captured the main
dynamics of the series.

Looking at the ACF, we don't notice significant spikes - indicating that
there is no remaining autocorrelations. This suggests that the residuals
behave like white noise.

This is also confirmed by the result of the Box-Ljung test:

-   **p-value of x** \<0.05 –\> the Ljung-Box test rejects the null
    hypothesis of independence, confirming that the series is not white
    noise and contains temporal dependence.

-   **p-value of y** \>0.05 –\> the Ljung-Box test does not reject the
    null hypothesis, indicating that the residuals are consistent with
    white noise. This validates the fitted AR(2) model.

So eventually, since the residuals behave as white noise, we conclude
that the AR(2) model provides an adequate fit to the data.

Alternative method:

```{r}

h<-floor(log(n))
tsdiag(s,gof.lag=h)
```

The resulting plots confirm that the residuals behave as white noise. In
particular, the p-values of the Ljung-Box test are generally above 0.05,
indicating no significant autocorrelation remains.

**Conclusions:**

The residual analysis shows that the fitted AR(2) model captures well
the structure of the series. The residuals derived are randomly
distributed, and their autocorrelation function shows no significant
values. Furthermore, the Ljung-Box test rejects independence for the
original series but not for the residuals. This confirms that the
residuals behave as white noise and that the model provides a good fit.

## 4. Selection of the best AR(p) model

The following block selects the AR model, AR(1), AR(2) or AR(3), that
best fits the data x, using the Akaike method, which we will see this in
next lectures. The solution should be an AR(2).

```{r}
n<-1000
x<-arima.sim(list(ar = c(0.8897, -0.4858), sd = sqrt(0.1796)), n)
plot(x,type="l",main="Simulated AR(2) series, phi = (0.8897, -0.4858)",xlab="Time",ylab="Value")
ss1<-ar(x, order.max=3, method="mle")
ss1
```

As expected, even with the Akaike method the AR(2) is selected as best
model to fit x.

## 5. A complete analysis with real data

We are going to use a dataset of the yearly level of Lake Huron from
years 1875 to 1972.

The exercise suggests to create the timseries and plot it, along with
its lag plots.

These are plots of (x(i-1), x(i)), then (x(i-2), x(i)), and so on. We
are asked to notice that the linear trend disappears as we increase the
lag.

```{r}
x<-LakeHuron
x

x <- LakeHuron

plot(x,main = "Lake Huron Water Levels (1875–1972)",xlab = "Year",ylab = "Depth")

lag.plot (x, lag=4, do.lines=F)
```

The time series represents yearly water levels of Lake Huron. The series
appears relatively stable, although a slight downward trend can be
observed over time, suggesting possible non-stationarity.

The **lag plots** show the relationship between the series and its past
values. For small lags, there is a clear linear relationship, indicating
strong dependence between consecutive observations. As the lag
increases, the relationship becomes more dispersed, suggesting
decreasing dependence.

```{r}
acf(x)
```

The autocorrelation function shows a slow decay, indicating strong
persistence in the series. This behavior suggests that the data may
follow an autoregressive process and may not be purely random.

Now fit a model AR(1) to x, using Yule-Walker & Maximum Likelihood
Estimation methods to estimate the parameters:

```{r}

s1<-ar(x,order.max=1, method="yw")
s1

s2<-ar(x,order.max=1, method="mle")
s2
```

Fitting an AR(1) model using both Yule-Walker and Maximum Likelihood
methods gives similar coefficient estimates. This indicates that both
methods provide consistent results for this dataset.

### Fit a generic AR(p) model

```{r}
s3 <- ar(x, method = "mle")
s3

s4 <- ar(x, method = "yw")
s4
```

Also in this case we can notice that both MLE and YW give similar
coefficient estimates. And they both select p=2, so they return AR(2)
models. This indicates robustness in the model selection and estimation
process.

**Conclusions:**

The analysis suggests that the Lake Huron time series exhibits
autoregressive behavior, with an AR(2) model providing a better fit than
AR(1). This is indeed coherent with what we saw in the ACG and lag
plots: they both suggested that the data was not purely random but
acutally followed an autoregressive process.

# Practical 3b Simulating and fitting ARMA models

## 1. Model simulation

```{r}
# Simulate ARMA(1,0), ARMA(0,2) and ARMA(2,1) models
SIM1 <- arima.sim(list(order=c(1, 0, 0), ar=0.7), n=1000)
SIM2 <- arima.sim(list(order=c(0, 0, 2), ma=c(1.5, 0.75)), n=1000)
SIM3 <- arima.sim(list(order=c(2, 0, 1), ar=c(0.3, 0.2), ma=0.3), n=1000)

# Plot the simulated series
par(mfrow=c(1,3))
plot(SIM1, type="l", main="SIM1: AR(1)")
plot(SIM2, type="l", main="SIM2: MA(2)")
plot(SIM3, type="l", main="SIM3: ARMA(2,1)")
```

**Interpretation**:

-   The AR(1) model (SIM1) shows an autocorrelation pattern, with values
    that tend to persist over time. Values depend on their immediate
    past, leading to smoother persistence.

-   The MA(2) model (SIM2) exhibits more "spiky" behavior, with short
    memory.

-   The ARMA(2,1) model (SIM3) combines features of both AR and MA,
    showing some persistence but also more variability than the pure
    AR(1) model.

## 2. Heuristic identification of the model. Auto-correlation and partial auto-correlation functions

```{r}
# Compute and plot the ACF and PACF for each simulated series
par(mfrow=c(3,2))

acf(SIM1, lag.max=30)
pacf(SIM1, lag.max=30)

acf(SIM2, lag.max=30)
pacf(SIM2, lag.max=30)

acf(SIM3, lag.max=30)
pacf(SIM3, lag.max=30)
```

**Interpretation**:

-   For SIM1 (AR(1)), the ACF shows a gradual decay, while the PACF has
    a significant spike at lag 1 and then drops off.

-   For SIM2 (MA(2)), the ACF has significant spikes at lags 1 and 2,
    while the PACF shows a gradual decay.

-   For SIM3 (ARMA(2,1)), the ACF and PACF show a more complex pattern,
    with both decay and significant spikes, reflecting the combination
    of AR and MA components.

## 3. Fitting a model

```{r}
# Fit ARMA models to the simulated data using maximum likelihood estimation
S1 <- arima(SIM1, order=c(1,0,0), method="ML")
S2 <- arima(SIM2, order=c(0,0,2), method="ML")
S3 <- arima(SIM3, order=c(2,0,1), method="ML")

# Display the fitted model summaries
S1
S2
S3
```

**Interpretation**:

-   The fitted AR(1) model for SIM1 show an estimated AR coefficient
    close to 0. (0.7172), confirming the true model.

-   The fitted MA(2) model for SIM2 show estimated MA coefficients close
    to 1.5 (1.5178) and 0.75 (0.7475), confirming the true model.

-   The fitted ARMA(2,1) model for SIM3 does not show estimated
    coefficients close to the true values (0.3, 0.2 for AR and 0.3 for
    MA). Obtained estimates are (AR(1) = 1.2707, AR(2) = −0.4478 and
    MA(1) = −0.6156).

## 4. Validation

```{r}
n <- length(SIM3)
h <- log(n)

Box.test(SIM3, lag=h, type="Ljung-Box")   # raw series
y <- resid(S3)
Box.test(y, lag=h, type="Ljung-Box")      # residuals
```

**Interpretation**:

-   The Ljung-Box test on the raw series (SIM3) yields a p-value of
    2.2e-16, rejecting the null hypothesis of independence, which is
    expected since SIM3 is an ARMA(2,1) model with autocorrelation and
    is not white noise.

-   The Ljung-Box test on the residuals of the fitted model (S3) yields
    a p-value of 0.3582 which is greater than the common significance
    level of 0.05, failing to reject the null hypothesis of
    independence, indicating that the residuals behave like white noise
    and suggesting that the fitted ARMA(2,1) model adequately captures
    the autocorrelation structure of the data.

### 4.1. Faster method

```{r}
# Diagnostic plots for the fitted ARMA(2,1) model
tsdiag(S3, gof.lag=20)
```

**Interpretation**:

-   The `tsdiag` function provides diagnostic plots for the fitted ARMA
    model (S3).

-   The first plot shows the standardized residuals, which should
    ideally resemble white noise.

-   The second plot displays the ACF of the residuals, where we look for
    no significant autocorrelations.

-   The third plot shows the p-values of the Ljung-Box test for
    different lags, where we want to see p-values above 0.05, indicating
    that the residuals are not significantly autocorrelated.

-   Overall, if the residuals appear random and the ACF shows no
    significant spikes, it suggests that the fitted ARMA(2,1) model is
    appropriate for the data.

## 5. A complete example

```{r}
# Simulate an ARMA(2,1) model with n=10000 observations
n <- 10000
h <- log(n)

# Simulate the ARMA(2,1) model
s3 <- arima.sim(list(order=c(2,0,1), ar=c(0.3,0.2), ma=0.3), n)

# Plot the simulated series and its ACF and PACF
plot(s3, type="l")
acf(s3)
pacf(s3)

# Fit the ARMA(2,1) model to the simulated data
p3 <- arima(s3, order=c(2,0,1), method="ML")
p3

Box.test(s3, lag=h, type="Ljung-Box")   # raw series
# Perform the Ljung-Box test on the residuals of the fitted model
Box.test(p3$residuals, lag=h, type="Ljung-Box")
tsdiag(p3, gof.lag=20)
```

**Interpretation**:

-   As said before, the Ljung-Box test on the raw series (s3) yields a
    p-value of 2.2e-16, rejecting the null hypothesis of independence

-   The Ljung-Box test on the residuals of the fitted model (p3) yields
    a p-value of 0.8576, which is greater than the common significance
    level of 0.05, failing to reject the null hypothesis of
    independence, indicating that the residuals behave like white noise
    and suggesting that the fitted ARMA(2,1) model adequately captures
    the autocorrelation structure of the data.

## 6. Roots of Polynomials

Given a polynomial 1+a_1 z + a_2 z\^2 + ... + a_p z\^p, we can obtain
its roots using the `polyroot` function in R. For an AR(2) model with
parameters (0.8897, -0.4858), we can compute the roots as follows:

```{r}
# Coefficients of the polynomial
a <- c(1, -0.8897, 0.4858)
# Compute the roots
roots <- polyroot(a)
roots
```

## 7. Download the daily closing prices for Siemens Energy and store them to a SE.csv file

```{python}
# import libraries
import yfinance as yf
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.dates as mdates
import requests
from statsmodels.tsa.stattools import adfuller
from statsmodels.graphics.tsaplots import plot_acf, plot_pacf
from statsmodels.tsa.arima.model import ARIMA
from statsmodels.stats.diagnostic import acorr_ljungbox
from statsmodels.tsa.statespace.sarimax import SARIMAX
from datetime import datetime
import warnings
warnings.filterwarnings("ignore")
```

### 7.1. Download the data

```{python}
ticker = "ENR.DE"  # Siemens Energy ticker on XETRA
data = yf.download(ticker, start="2020-01-01", end="2025-12-31")
data.to_csv("SE.csv")  # save to CSV
print(data.head())
```

### 7.2. Load the CSV and compute log returns

```{python}
#  yfinance has extra row at the beginning, skip it with header=1
se_data = pd.read_csv("SE.csv", header=1, index_col=0, parse_dates=True)
se_data.columns = se_data.columns.str.strip()
print(se_data.columns)

# Compute daily log returns
# log_return[†] = log(Close_t) - log(Close_{t-1})
returns = np.log(se_data['ENR.DE.3']).diff().dropna()

# Plot
returns.plot(title="Siemens Energy Daily Log Returns", figsize=(10,6))
plt.show()
```

### 7.3. Check stationarity with ADF test

```{python}
adf_result = adfuller(returns)
print(f"ADF Statistic: {adf_result[0]}")
print(f"p-value: {adf_result[1]}")
print("Critical Values:", adf_result[4])
```

**Interpretation**:

-   The log returns are stationary because the ADF test statistic is
    less than the critical value and the p-value is less than 0.05

-   ADF Statistic: -40.30339446788923

-   p-value: 0.0

-   We can reject the null hypothesis of a unit root, confirming that
    the log returns are stationary.

### 7.4. Plot ACF and PACF

```{python}
# Plot ACF
plt.figure(figsize=(12,5))
plot_acf(returns, lags=50, zero=False)
plt.title("ACF of Siemens Energy Log Returns")
plt.show()

# Plot PACF
plt.figure(figsize=(12,5))
plot_pacf(returns, lags=50, zero=False, method='ywm')  # 'ywm' handles large series better
plt.title("PACF of Siemens Energy Log Returns")
plt.show()
```

### 7.5. Fit an ARMA(1, 1) and ARMA(2, 1) model based on ACF and PACF

```{python}
# ARMA(p, q) is ARIMA(p, 0, q) in statsmodels
model_11 = ARIMA(returns, order=(1, 0, 1)) # p=1, d=0, q=1
fit_11 = model_11.fit()
print(fit_11.summary())

model_21 = ARIMA(returns, order=(2, 0, 1)) # p=2, d=0, q=1
fit_21 = model_21.fit()
print(fit_21.summary())
```

### Interpret the AR and MA coefficients. What do they say about short-term dependence in returns?

Let's first interpret the ARMA(1, 1) coefficients. From the results:

-   AR(1) = -0.4588
-   MA(1) = 0.3663
-   σ² aprox 0.001

The **autoregressive term (AR)** measures how much yesterday's return
influences today's return. Negative value implies mean-reversion. If
yesterday's return was positive, today's return is slightly pulled
downward and if yesterday's return was negative, today's return tends to
move upwards. This indicates that returns slightly reverse previous
day's movements. The **moving average (MA)** term measures how much
yesterday's stock (unexpected return) affects today's return. Positive
value implies shock persistence. If yesterday unexpected return was
positive, today return will slightly increase and viceversa. This
indicates that short-term surprises in returns have a small effect on
the next day.

**Conclusion:**

-   Negative AR(1) -\> returns tend to bounce back

-   Positive MA(1) -\> short-lived stock persistence, sudden movements
    carry over to the next day.

-   Daily log returns are weakly autocorrelated. Past values and past
    stocks have a small but measurable impact.

For ARMA(2, 1) coefficients:

-   AR(1) = -0.8779 (very large standard error, not significant)
-   AR(2) = -0.0755 (very large standard error, not significant)
-   MA(1) = 0.7802 (very large standard error, not significant)
-   σ² aprox 0.001

The **autoregressive terms (AR1, AR2)** are not statistically
significant, they have huge standard errors, so estimates are
unreliable. Negative AR1 hints at mean-reversion, but due to
insignificance we cannot confidently say returns show short-term
dependence based on just AR terms. AR2 is close to zero -\> little
effect of returns from two years ago. The **moving average (MA1)** term
is positive, but again not significant. This suggest that past stocks
may have some short-lived effect, but we cannot rely on the estimate.

**Conclusion:**

-   Model does not improve significantly over ARMA(1, 1). None of the
    additional parameters (AR2, MA1) are significant.

-   Short term dependence is captured by the ARMA(1, 1) model.

-   Adding a second AR term doesn't add meaningful predictions.

### 7.6. Forecast next 5 days of returns and visualize them vs historical returns

```{python}
# forecast 5 days
forecast_steps = 5
forecast = fit_11.get_forecast(steps=forecast_steps)
forecast_mean = forecast.predicted_mean
forecast_ci = forecast.conf_int()

# returns.index is datetime
returns.index = pd.to_datetime(returns.index)

# create forecast index 5 business days
last_date = returns.index[-1]
forecast_index = pd.bdate_range(start=last_date + pd.Timedelta(days=1), periods=forecast_steps)

# plot historical and forecasted returns
plt.figure(figsize=(12,6))
plt.plot(returns.index, returns, label="Historical Returns", color='blue')
plt.plot(forecast_index, forecast_mean, label="Forecasted Returns (5 days)", color='red', marker='o')
plt.fill_between(forecast_index, forecast_ci.iloc[:,0], forecast_ci.iloc[:,1], color='pink', alpha=0.3)

plt.title("Siemens Energy Daily Log Returns: Historical vs 5-Day Forecast")
plt.xlabel("Date")
plt.ylabel("Log Return")
plt.legend()
plt.show()
```

### 7.7. Compare the performance of ARMA(1,1) vs ARMA(2,1) using AIC/BIC

```{python}
# ARMA(1,1) summary
print("ARMA(1,1) AIC:", fit_11.aic)
print("ARMA(1,1) BIC:", fit_11.bic)

# ARMA(2,1) summary
print("ARMA(2,1) AIC:", fit_21.aic)
print("ARMA(2,1) BIC:", fit_21.bic)
```

**Interpretation:**

-   Both AIC and BIC are lower for ARMA(1, 1), meaning it fits the data
    better while being simpler.

-   ARMA(2, 1) adds extra parameters that do not significantly improve
    the fit.

-   Conclusion: ARMA(1, 1) is preferred over ARMA(2, 1) for Siemens
    Energy daily log returns; it captures short-term dependency without
    unnecessary complexity.

## 8. ARMA for weather data

### 8.1. Load the csv file and visualize the time series. Is there any visible trend or seasonality?

We use the free [Open-Meteo Historical Weather
API](https://open-meteo.com/) to download daily mean temperatures for
**Barcelona, Spain** (lat 41.39, lon 2.15) for 2022–2025.

```{python}
url = "https://archive-api.open-meteo.com/v1/archive"
params = {
    "latitude":        41.3888,
    "longitude":       2.1590,
    "start_date":      "2020-01-01",
    "end_date":        "2025-12-31",
    "daily":           "temperature_2m_mean",
    "timezone":        "Europe/Madrid",
}

response = requests.get(url, params=params, timeout=30)
response.raise_for_status()
data = response.json()

df = pd.DataFrame({
    "date": pd.to_datetime(data["daily"]["time"]),
    "temp": data["daily"]["temperature_2m_mean"],
})
df = df.dropna().reset_index(drop=True)
df = df.set_index("date")

print(f"Fetched {len(df)} daily observations: {df.index[0].date()} → {df.index[-1].date()}")
print(df.describe().round(2))
```

### 8.2. Visualize the raw time series

```{python}
fig, ax = plt.subplots(figsize=(12, 4))
ax.plot(df.index, df["temp"], color="#3266ad", linewidth=0.9, alpha=0.85)
ax.set_title("Barcelona Daily Mean Temperature (2020–2025)")
ax.set_ylabel("Temperature (°C)")
ax.xaxis.set_major_formatter(mdates.DateFormatter("%b %Y"))
ax.xaxis.set_major_locator(mdates.MonthLocator(interval=3))
plt.xticks(rotation=30)
plt.tight_layout()
plt.show()
```

**Interpretation:** We see a clear seasonality (aprox. 10 degrees in
winter, aprox. 25 degrees in summer). There's no obvious long-term
trend. The series is non-stationary due to seasonality.

### 8.3. Stationary test (ADF)

```{python}
adf_result = adfuller(df["temp"], autolag="AIC")
print(f"ADF Statistic : {adf_result[0]}")
print(f"p-value       : {adf_result[1]}")
print(f"Critical (5%) : {adf_result[4]['5%']}")
```

**Interpretation:** The ADF test on the raw series shows p = 0.091, not
stationary at 5% but borderline. This is expected because the series has
seasonal non-stationarity (annual cycle), not just a trend.

### 8.4. Difference the series to achieve stationarity

We apply first-order differencing to remove the seasonal mean and make
the series stationary.

```{python}
diff = df["temp"].diff().dropna()

fig, axes = plt.subplots(2, 1, figsize=(12, 6), sharex=True)

axes[0].plot(df.index, df["temp"], color="#3266ad", linewidth=0.8)
axes[0].set_title("Original series")
axes[0].set_ylabel("Temp (°C)")

axes[1].plot(diff.index, diff.values, color="#D85A30", linewidth=0.8)
axes[1].axhline(0, color="black", linewidth=0.8, linestyle="--")
axes[1].set_title("First-differenced series  dy_t")
axes[1].set_ylabel("dTemp (°C)")
axes[1].xaxis.set_major_formatter(mdates.DateFormatter("%b %Y"))
axes[1].xaxis.set_major_locator(mdates.MonthLocator(interval=3))

plt.xticks(rotation=30)
plt.tight_layout()
plt.show()

# ADF on differenced series
adf2 = adfuller(diff, autolag="AIC")
print(f"ADF on differenced: statistic={adf2[0]:.4f}, p-value={adf2[1]:.6f}")
```

We obtain a stationary series as p-value=0.000000

### 8.5. ACF and PACF of the Differenced Series

```{python}
# Use only lag-1 differenced series (from the plot above)
diff = df["temp"].diff().dropna()

fig, axes = plt.subplots(1, 2, figsize=(14, 4))

plot_acf(diff, lags=40, ax=axes[0], color="#3266ad",
         title="ACF — First-differenced series", zero=False)
plot_pacf(diff, lags=40, ax=axes[1], color="#D85A30",
          title="PACF — First-differenced series", zero=False, method="ywm")

for ax in axes:
    ax.set_xlabel("Lag (days)")
    ax.axhline(0, color="black", linewidth=0.5)
    # Add reference lines at lag 7 and 365 to spot weekly/annual patterns
    for lag in [7, 14, 21]:
        ax.axvline(lag, color="gray", linestyle=":", linewidth=0.7, alpha=0.5)

plt.tight_layout()
plt.show()
```

Nearly all lags within the confidence bands. The first-differenced
Barcelona temperature series behaves like random walk with noise. Daily
temperatures changes are unpredictable from past changes alone. This is
realistic, as whether tomorrow is 1 degree warmer or cooler than today
carries almost no information about the day after. The only signal
perceived is a small negative cluster at lags 1-5 in the PACF (values
around -0.10 to -0.15), which barely touched the band edge. This is the
only exploitable structure from the plots.

**Model selection**: we will select ARIMA (1, 1, 1). We don't want to
overfit with AR(2) or higher.

### 8.6. Fit ARIMA(1, 1, 1) Model

```{python}
model = ARIMA(df["temp"], order=(1, 1, 1))
result = model.fit()
print(result.summary())

phi1   = result.arparams[0]
theta1 = result.maparams[0]
sigma2 = result.params["sigma2"]
print(f"\nφ₁  = {phi1:.4f}   (AR lag-1)")
print(f"θ₁  = {theta1:.4f}   (MA lag-1)")
print(f"σ²  = {sigma2:.4f}")
```

```{python}
# residual diagnostics plot
resid = result.resid.dropna()

fig, axes = plt.subplots(1, 3, figsize=(15, 4))

# residuals over time
axes[0].plot(resid.index, resid.values, color="#888780", linewidth=0.6)
axes[0].axhline(0, color="black", linewidth=0.8, linestyle="--")
axes[0].set_title("Residuals over time")
axes[0].set_ylabel("ε_t (°C)")

# distribution
axes[1].hist(resid, bins=50, color="#3266ad", edgecolor="white", linewidth=0.3)
axes[1].set_title(f"Residual distribution\nKurtosis = 4.30  (Normal = 3.0)")

# ACF of residuals
from statsmodels.graphics.tsaplots import plot_acf
plot_acf(resid, lags=20, ax=axes[2], color="#D85A30",
         title="ACF of residuals\n(should be white noise)", zero=False)

plt.tight_layout()
plt.show()
```

The model is well specified, the only real issue is non-gaussianity...

### 8.7. Forecast next 7 days of temperature changes

```{python}
steps = 7 # days
forecast = result.get_forecast(steps=steps)
fc_level = forecast.predicted_mean # temperature level
fc_ci = forecast.conf_int(alpha=0.05)

last_date = df.index[-1]
last_temp = df["temp"].iloc[-1]
fc_dates  = pd.date_range(last_date + pd.Timedelta(days=1), periods=steps, freq="D")
fc_level.index = fc_dates
fc_ci.index    = fc_dates

prev_levels = np.concatenate([[last_temp], fc_level.values[:-1]])
delta_hat   = fc_level.values - prev_levels
```

### 8.8. Convert the predicted changes back to temperature predictions + visualize

```{python}
cumulative = np.cumsum(delta_hat)
manual_temps = last_temp + cumulative # should match fc_level.values exactly
print(np.round(manual_temps - fc_level.values, 6)) # should be all aprox 0
```

### 8.8. Plot actual vs fitted values for the last 30 days

```{python}
fitted_vals = result.fittedvalues
last30_actual = df["temp"].iloc[-30:]
last30_fitted = fitted_vals.loc[last30_actual.index]
residuals_30  = last30_actual - last30_fitted

fig, ax = plt.subplots(figsize=(13, 5))

ax.plot(last30_actual.index, last30_actual.values,
        color="#3266ad", linewidth=2, marker="o",
        markersize=4, label="Actual T_t")
ax.plot(last30_fitted.index, last30_fitted.values,
        color="#D85A30", linewidth=1.8, linestyle="--",
        marker="s", markersize=4, label="Fitted T̂_t  ARIMA(1,1,1)")
ax.fill_between(last30_actual.index,
                last30_actual.values, last30_fitted.values,
                alpha=0.12, color="#888780", label="Error region")
ax.set_title("actual vs fitted last 30 days", fontweight="bold")
ax.set_ylabel("Temperature (deg)")
ax.legend()
ax.xaxis.set_major_formatter(mdates.DateFormatter("%d %b"))
plt.xticks(rotation=25)
plt.tight_layout()
plt.show()

rmse = np.sqrt(np.mean(residuals_30**2))
mae  = np.mean(np.abs(residuals_30))
print(f"RMSE : {rmse:.3f} degrees")
print(f"MAE  : {mae:.3f} degrees")
```

The fitted line tracks the actual one with a one-day lag. The model
always knows where temperature was yesterday (AR) and corrects for
yesterday's error (MA). It follows the shape well but misses turning
points by one step. Limitation of ARIMA model: cannot anticipate
direction changes, only follow them.

The gray area (error) is narrow most of the time, widening on 04 Dec and
24 Dec, corresponding to extreme days that caused the leptokurdic
residual distribution saw in previous plots.

### 8.9. Optional: try using a seasonal ARIMA (SARIMA) if there is weekly or yearly seasonality in temperature

Daily temperature has annual seasonality (summer/winter cycle), not
weekly. The seasonal period we will use is m=365.

```{python}
sarima = SARIMAX(
    df["temp"],
    order=(1, 1, 1),
    seasonal_order=(1, 0, 0, 365),
    enforce_stationarity=False,
    enforce_invertibility=False,
)
sarima_result = sarima.fit(disp=False)
print(sarima_result.summary())
```

Why (1, 0, 0)[365] seasonal part? - P=1 -\> AR at lag 365: today
correlates with same day last year - D=0 -\> no seasonal differencing
(d=1 already handles stationarity) - Q=0 -\> no seasonal MA

**Parameter interpretation:**

```{python}
phi1_s  = sarima_result.arparams[0]       # non-seasonal AR(1)
theta1_s = sarima_result.maparams[0]      # non-seasonal MA(1)
Phi1    = sarima_result.seasonalarparams[0]  # seasonal AR(1) at lag 365
print(f"\nNon-seasonal  φ₁  = {phi1_s:.4f}")
print(f"Non-seasonal  θ₁  = {theta1_s:.4f}")
print(f"Seasonal      Φ₁  = {Phi1:.4f}")
```

-   φ₁ = 0.7271 -\> today's temperature change is 73% correlated with
    yesterday's
-   θ₁ = -0.8666 -\> the model rapidly undoes yesterday's forecast
    mistake
-   Φ₁ = 0.0202 -\> the seasonal correction from 365 days ago
    contributes only 2% weight... Practically zero.

This means that the 7-day forecast from SARIMA will be almost identical
to the plain ARIMA one, because the seasonal term contributes the most.

**Model comparison:**

```{python}
print(f"ARIMA(1,1,1): AIC: {result.aic:.1f}, BIC: {result.bic:.1f}")
print(f"SARIMA(1,1,1)(1,0,0)[365]: AIC: {sarima_result.aic:.1f}, BIC: {sarima_result.bic:.1f}")
delta_aic = result.aic - sarima_result.aic
```

**Fitted vs actual last 30 days: ARIMA vs SARIMA**

```{python}
sarima_fitted = sarima_result.fittedvalues
s_last30 = sarima_fitted.loc[last30_actual.index]
s_residuals_30 = last30_actual - s_last30

fig, ax = plt.subplots(figsize=(13, 5))

ax.plot(last30_actual.index, last30_actual.values,
        color="#3266ad", linewidth=2, marker="o",
        markersize=4, label="Actual T_t")
ax.plot(last30_fitted.index, last30_fitted.values,
        color="#D85A30", linewidth=1.6, linestyle="--",
        marker="s", markersize=3, label="ARIMA(1,1,1) fitted", alpha=0.8)
ax.plot(s_last30.index, s_last30.values,
        color="#639922", linewidth=1.6, linestyle="-.",
        marker="^", markersize=4, label="SARIMA(1,1,1)(1,0,0)[365] fitted")

ax.set_title("Actual vs Fitted last 30 days\nARIMA vs SARIMA")
ax.set_ylabel("Temperature (deg)")
ax.legend()
ax.xaxis.set_major_formatter(mdates.DateFormatter("%d %b"))
plt.xticks(rotation=25)
plt.tight_layout()
plt.show()

# Metrics comparison
rmse_s = np.sqrt(np.mean(s_residuals_30**2))
mae_s  = np.mean(np.abs(s_residuals_30))
print(f"\n Last-30-day comparison")
print(f"ARIMA  RMSE: {rmse:.3f} degrees   MAE: {mae:.3f} degrees")
print(f"SARIMA RMSE: {rmse_s:.3f} degrees   MAE: {mae_s:.3f} degrees")
print(f"RMSE improvement: {rmse - rmse_s:.3f} degrees")
```
