---
title: "R Notebook"
output: html_notebook
---

# Practical 3a AR(p) Models

## 1. Simulating AR(p) models
Generate an AR(2) model:

```{r}

n<-1000
x<-arima.sim(list(ar = c(0.8897, -0.4858), sd = sqrt(0.1796)),n)
plot(x,type="l")

```

Compute its autocorrelation function:

```{r}
acf(x)
```

Repeat the same exercise with different values of n, with a different number of parameters, and with different parameters of the AR(p).

First I create a function to build the AR(p) model, so that I can easily vary n, p and the parameters phi:
```{r}
simulate_ar <- function(n, phi, sigma2 = 1) {
  
  x <- arima.sim(
    model = list(ar = phi),
    n = n,
    sd = sqrt(sigma2)
  )
  
  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par))
  
  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  
  plot(
    x,
    type = "l",
    main = paste0("AR(", length(phi), "), n = ", n),
    xlab = "Time",
    ylab = "Value"
  )
  
  acf(x, main = "ACF")
}
```

### REPEAT FOR DIFFERENT n
```{r}
n_values <- c(100, 500, 1000)

for (n in n_values) {
  simulate_ar(n, phi = c(0.8897, -0.4858), sigma2 = 0.1796)
  dev.flush()
}
```

### REPEAT FOR DIFFERENT p (AR orders)
```{r}
phi_list <- list(
  c(0.7), #AR(1) 
  c(0.8897, -0.4858), #AR(2)
  c(0.6, -0.3, 0.2) #AR(3)
)

for (phi in phi_list) {
  simulate_ar(1000, phi)
}
```

### REPEAT FOR DIFFERENT n AND p COMBINED
```{r}
n_values <- c(100, 500, 1000)

phi_list <- list(
  c(0.7),
  c(0.8897, -0.4858),
  c(0.6, -0.3, 0.2)
)

for (n in n_values) {
  for (phi in phi_list) {
    
    x <- arima.sim(list(ar = phi), n = n, sd = sqrt(0.1796))
    
    par(mfrow=c(1,2))
    
    plot(x, type="l",
         main=paste("AR(",length(phi),")  n=",n,sep=""))
    
    acf(x, main="ACF")
    
  }
}
```

